Interacting with Modal Logics \\ in the Coq Proof Assistant

### 2014-12-11 Papers/InProgress/ModalLogicCoq/ModalCoq.tex
Improve language of sections 1 to 5

\title{
(Interacting with)? \\ Modal Logics in the Coq Proof Assistant
}

\author{
  Christoph Benzm\"{u}ller\inst{1}\thanks{This work has been supported by the German Research Foundation (DFG) under grant BE2501/9-1.} 
  \and 
  Bruno Woltzenlogel Paleo\inst{2}
}

@@

Modal logics extend usual formal logic languages by adding modal
operators ($\Box$ and $\Diamond$) and are characterized by the
\emph{necessitation rule}, according to which $\Box A$ is a theorem if
$A$ is a theorem, although $A \imp \Box A$ is not necessarily a
theorem. Various informal concepts, such as \emph{necessity and
possibility}, \emph{knowledge and belief}, \emph and \emph{temporal
globality and eventuality}, have been formalized with the help of
modal operators. Therefore, modal logics \cite{ModalLogic} are well-
established for a wide variety of application domains, ranging from
philosophy \cite{ToDo} to formal software verification
\cite{ToDo:TLAPlus}.\footnote{ChrisToBruno: \cite{ModalLogic} should
  suffice as reference, applications are addressed in Part 4.}

However, general automated reasoning support for modal logics is still
not as well-developed as for classical logics. Deduction tools
for modal logics are often limited to propositional, quantifier-free,
fragments \cite{ToDo:SeeBjornsProposal} or tailored to particular
modal logics and their applications \cite{ToDo:StephanMerz}; 
first-order automated deduction techniques based on tableaux, sequent
calculi and connection calculi have only recently been generalized 
and implemented in a few new provers able to directly cope with
modalities \cite{JensOtten}.

Another recently explored possibility \cite{J23,B9,J22} is the
embedding of first-order and even higher-order modal logics into classical
higher-order logics, for which existing higher-order automated theorem
provers \cite{LEO-II,Satallax} exist.  Embedding is flexible,
because various modal logics can be easily supported by adding their
characteristic axioms.  The embedding approach can also be easily
adapted to support multiple modalities, and it is possible to replace
constant domain quantifiers by varying or cumulative domain
quantifiers.  Moreover, the approach is relatively simple to
implement, because it does not require any modification in the source
code of the higher-order prover. The prover can be used as is, and
only the input files provided to the prover must be especially
encoded. Furthermore, the efficacy and efficiency of the embedding
approach has been confirmed in philosophically interesting and
relevant benchmarks \cite{FormalTheologyRepository}. These qualities
make embedding a convenient approach for \emph{fully automated}
reasoning.

However, one may wonder whether the embedding approach is adequate
also for \emph{interactive} reasoning, when the user proves theorems
by interacting with a proof assistant such as \Coq\footnote{The \Coq
proof assistant was chosen merely because of the authors' greater
familiarity with the tactic language of this system. Nevertheless, the
techniques presented and discussed here are likely to be useful for
other proof assistants, such as \texttt{Isabelle} \cite{Isabelle} or
\texttt{HOL-Light} \cite{HOLLight}, as well.}. The main goal of this
paper is to study this question. Our answer is positive.

@@

Propositional and quantified modal logics have applications in various
fields, including, for example, verification, agent technologies,
artificial intelligence, computational linguistics, and computational
philosophy. Our contribution may thus serve as a starting point for
many interesting further projects.

@@

The simpler ontological argument of Anselm has been automatically
verified with \texttt{PVS} by Rushby \cite{Rushby} and with first-
order theorem provers by Oppenheimer and Zalta \cite{Zalta}. Gödel's
argument was automatically verified in our previous work on embedding-
based fully automated modal theorem proving \cite{arXiv,AFP}. This
paper presents the first fully interactive and detailed formalization
of this proof in a proof assistant. Hence we here reveal the proof
structure on a cognitively adequate level of
detail. Such detailed proof structure has been hidden in our
other papers on the subject due to the use of automated theorem
provers. In comparison to \cite{Rushby,Zalta}, however, our
contribution stands out with its surprising technical 
simplicity and elegance.

@@

A crucial aspect of modal logics \cite{ModalLogic} is that the so-
called \emph{necessitation rule} allows $\Box A$ to be derived if $A$
is a theorem, but $A \imp \Box A$ is not necessarily a theorem. Naive
attempts to define the modal operators $\Box$ and $\Diamond$ may
easily be unsound in this respect. To avoid this issue, the
\emph{possible world semantics} of modal logics can be explicitly
embedded into higher-order logics \cite{J23,B9}.

The embedding is related to labeling techniques \cite{Labels}.
However, the expressiveness of higher-order logic is exploited here to
encode the labels within the logical language. To this aim, a type for
worlds must be declared and modal propositions should be not of type
\texttt{Prop} but of a lifted type \texttt{o} that depends on possible
worlds:

@@

Interactive theorem proving in \Coq is usually done with tactics,
imperative commands that reduce the theorem to be proven (i.e. the
goal) to simpler subgoals, in a bottom-up manner. The simplest tactics
can be regarded as rules of a natural deduction calculus\footnote{The
underlying proof system of \Coq (the Calculus of Inductive
Constructions \cite{Paulin}) is actually more sophisticated and
minimalistic than the calculus shown in Figure
\ref{fig:NaturalDeduction}. But the calculus shown here suffices for
the purposes of this paper.} \footnote{ChrisToBruno: We need to briefly address here
    the fact that this calculus is classical while Coq pure is
    constructive; one or two sentences should be fine.} 
(e.g. as those shown in Figure
\ref{fig:NaturalDeduction}).  For example: the \texttt{intro} tactic
can be used to apply the introduction rules for implication and for
the universal quantifier; the \texttt{apply} tactic corresponds to the
elimination rules for implication and for the universal quantifier;
\texttt{split} performs conjunction introduction; \texttt{exists} can
be used for existential quantifier introduction and \texttt{destruct}
for its elimination.

@@

Since the modal operators are, in our embedding, essentially just
abbreviations for quantifiers guarded by reachability conditions, the
typical tactics for quantifiers can be used, in principle.  However,
this exposes the user to the technicalities of the embedding,
requiring him to deal with possible worlds and their reachability
explicitly. In order to obtain transparency also for the modal
operators, we have implemented a few specialized tactics using \Coq's
Ltac language. We present these tactics next.

@@

Finally, our implementation also provides the tactic \texttt{mv},
standing for \emph{modal validity}, which replaces a goal of the form
\texttt{[ p ]} (or equivalently \texttt{(V p)}) by a goal of the form
\texttt{(p w)} for a fresh arbitrary world \texttt{w}.

@@

\noindent
For other modal logics beyond \textbf{K}, their frame conditions constraining 
the reachability relation must be stated as \Coq axioms.

@@

\noindent
In strong modal logics, such as S5 (which enforces all three frame conditions specified above), iterations of modal operators can be collapsed. This is a controversial principle used in G\"odel's proof shown in Section \ref{sec:proof}.

@@

\noindent 
The second part of the proof consists in showing that if
God's existence is possible then it must be necessary
(\texttt{lemma2}), for which the \textbf{S5} principle
\texttt{dia\_box\_to\_box} is used.

>>>>

\title{
Interacting with Modal Logics \\
in the Coq Proof Assistant
}

\author{
  Christoph Benzm\"{u}ller\inst{1}\thanks{Supported by the German Research Foundation (DFG) under grant BE2501/9-1.} 
  \and 
  Bruno Woltzenlogel Paleo\inst{2}
}

@@

Modal logics \cite{ModalLogic} extend usual formal logic languages by adding modal
operators ($\Box$ and $\Diamond$) and are characterized by the
\emph{necessitation rule}, according to which $\Box A$ is a theorem if
$A$ is a theorem, even though $A \imp \Box A$ is not necessarily a
theorem. Various notions, such as \emph{necessity and
possibility}, \emph{obligation and permission}, 
\emph{knowledge and belief}, \emph and \emph{temporal
globality and eventuality}, which are ubiquitous in various application domains,
have been formalized with the help of modal operators.

Nevertheless, general automated reasoning support for modal logics is still
not as well-developed as for classical logics. Deduction tools
for modal logics are often limited to propositional, quantifier-free,
fragments \cite{ToDo:SeeBjornsProposal} or tailored to particular
modal logics and their applications \cite{ToDo:StephanMerz}; 
first-order automated deduction techniques based on tableaux, sequent
calculi and connection calculi have only recently been generalized 
and implemented in a few new provers able to directly cope with
modalities \cite{JensOtten}.

Another recently explored possibility is the
embedding of first-order and even higher-order modal logics into classical
higher-order logics \cite{J23,B9,J22}, 
for which existing higher-order automated theorem
provers \cite{LEO-II,Satallax} exist. The embedding approach is flexible,
because various modal logics (even with multiple modalities or varying/cumulative domain quantifiers) can be easily supported by stating their
characteristic axioms. Moreover, the approach is relatively simple to
implement, because it does not require any modification in the source
code of the higher-order prover. The prover can be used as is, and
only the input files provided to the prover must be especially
encoded. Furthermore, the efficacy and efficiency of the embedding
approach has been confirmed in philosophically interesting and
relevant benchmarks \cite{FormalTheologyRepository}. These qualities
make embedding a convenient approach for \emph{fully automated}
reasoning.

However, one may wonder whether the embedding approach is adequate
also for \emph{interactive} reasoning, when the user proves theorems
by interacting with a proof assistant such as \Coq\footnote{The \Coq
proof assistant was chosen merely because of the authors' greater
familiarity with the tactic language of this system. Nevertheless, the
techniques presented and discussed here are likely to be useful for
other proof assistants, such as \texttt{Isabelle} \cite{Isabelle} or
\texttt{HOL-Light} \cite{HOLLight}, as well.}. The main goal and novelty
of this paper is to study this question. Our answer is positive.

@@

@@

The simpler ontological argument of Anselm has been automatically
verified with \texttt{PVS} by Rushby \cite{Rushby} and with 
first-order theorem provers by Oppenheimer and Zalta \cite{Zalta}. Gödel's
argument was automatically verified in our previous work on embedding-
based fully automated modal theorem proving \cite{arXiv,AFP}. This
paper presents the first fully interactive and detailed formalization
of this proof in a proof assistant. Hence we here reveal the proof
structure on a cognitively adequate level of
detail. Such detailed proof structure has been hidden in our
other papers on the subject due to the use of automated theorem
provers. In comparison to \cite{Rushby,Zalta}, however, our
contribution stands out with its surprising technical 
simplicity and elegance.

In addition to philosophy, propositional and quantified modal logics 
have (potential) applications in various other fields as well, including, for instance, 
verification, artificial intelligence
agent technologies, law and linguistics. \marginpar{BrunoToChris: Do we need citations for these applications?} 
Therefore, the main contribution described in this paper -- convenient techniques for leveraging a powerful proof assistant such as \Coq for interactive reasoning for modal logics -- may serve as a starting point for
many interesting projects.

@@

A crucial aspect of modal logics \cite{ModalLogic} is that the 
so-called \emph{necessitation rule} allows $\Box A$ to be derived if $A$
is a theorem, but $A \imp \Box A$ is not necessarily a theorem. Naive
attempts to define the modal operators $\Box$ and $\Diamond$ may
easily be unsound in this respect. To avoid this issue, the
\emph{possible world semantics} of modal logics can be explicitly
embedded into higher-order logics \cite{J23,B9}.

The embedding technique described in this section is 
related to labeling techniques \cite{Labels}.
However, the expressiveness of higher-order logic can be exploited in order to
encode the labels within the logical language itself. To this aim, a type for
worlds must be declared and modal propositions should be not of type
\texttt{Prop} but of a lifted type \texttt{o} that depends on possible
worlds:

@@

Interactive theorem proving in \Coq is usually done with tactics,
imperative commands that reduce the theorem to be proven (i.e. the
goal) to simpler subgoals, in a bottom-up manner. The simplest tactics
can be regarded as rules of a natural deduction calculus\footnote{The
underlying proof system of \Coq (the Calculus of Inductive
Constructions (CIC) \cite{Paulin}) is actually more sophisticated and
minimalistic than the calculus shown in Figure
\ref{fig:NaturalDeduction}. But the calculus shown here suffices for
the purposes of this paper. This calculus is classical, because of the double negation elimination rule. Although CIC is intuitionistic, it can be made classical by importing \Coq's classical library, which adds the axiom of the \emph{excluded middle} and the double negation elimination lemma.}
(e.g. as those shown in Figure
\ref{fig:NaturalDeduction}).  For example: the \texttt{intro} tactic
can be used to apply the introduction rules for implication and for
the universal quantifier; the \texttt{apply} tactic corresponds to the
elimination rules for implication and for the universal quantifier;
\texttt{split} performs conjunction introduction; \texttt{exists} can
be used for existential quantifier introduction and \texttt{destruct}
for its elimination.

@@

Since the modal operators are, in our embedding, essentially just
abbreviations for quantifiers guarded by reachability conditions, the
typical tactics for quantifiers can be used, in principle.  However,
this exposes the user to the technicalities of the embedding,
requiring him to deal with possible worlds and their reachability
explicitly. In order to obtain transparency also for the modal
operators, we have implemented specialized tactics using \Coq's
Ltac language. These tactics are among our main contributions and they are described in the remainder of this section.

@@

\noindent
Finally, our implementation also provides the tactic \texttt{mv},
standing for \emph{modal validity}, which replaces a goal of the form
\texttt{[ p ]} (or equivalently \texttt{(V p)}) by a goal of the form
\texttt{(p w)} for a fresh arbitrary world \texttt{w}.

@@

\noindent
For other modal logics beyond \textbf{K}, their frame conditions, which constrain 
the reachability relation, must be stated as \Coq axioms.

@@

\noindent
In a strong modal logic such as \textbf{S5} (which requires all three frame conditions specified above), sequences of modal operators can be collapsed to a single modal operator. One such collapsing principle is specified and proven below. By applying it iteratively, any sequence $\pos \ldots \pos \nec p$ could be collapsed to $\nec p$.

@@

\noindent 
The second part of the proof consists in showing that if
God's existence is possible then it must be necessary
(\texttt{lemma2}), for which the controversial \textbf{S5} principle
\texttt{dia\_box\_to\_box} is used.
